Engineering the Servo Web Browser Engine using Rust

### 2015-10-24 icse16-servo/rust.tex
minor text and comment tweaks

Rust's most prominent influence from the ML languages are algebraic data types,
along with \emph{pattern matching} to destructure
them into their components. Rust calls algebraic data types \emph{enums},
and they are the union of multiple types,
each of which contains their own fields. Although superficially
similar to C unions, they are significantly different in that
instances of enums cannot be indiscriminately cast between
variants; to access the values in an enum one must employ the \lstinline[language=Rust]{match}
expression to first check the variant, then bind references
to its interior fields. In \figref{fig:enums} the values of
a drawing message are extracted from an enum for further processing,
with the bindings to the enum's fields established to the left of the
fat arrow (\lstinline[language=Rust]{=>}) and the block of code operating
on those bindings to the right.

@@

Both enums and structs (which are substantially similar to C structs)
may have associated instance methods and static methods, called with the dot
(\lstinline[language=Rust]{.}) operator and the double-colon
(\lstinline[language=Rust]{::}) operator, respectively.
Static methods are common in Rust as they are used to construct
values, conventionally with a method called
\lstinline[language=Rust]{new}, as in
\lstinline[language=Rust]{Point2D::new(0.0, 0.0)}.
Notably, methods in Rust are dispatched statically, making them
eligible for inlining and aggressive optimization; methods are never
dispatched dynamically through a function pointer in Rust unless
explicitly requested.

@@

Finally, Rust has a hygienic macro system, much more powerful than
the C preprocessor. Macro invocations look like function invocations
where the function name is appended with a bang (\lstinline[language=Rust]{!}).
They are not functions though, but are expanded in place at compile time,
and perform syntactic transformations unavailable to functions.
The common method of performing console output, \lstinline[language=Rust]~println!(``{}'', foo)~,
is a macro.
See \figref{fig:bad-concurrency} for an example of closures and macros.

\subsection{Ownership}
The features of Rust described so far are common to many languages.
Rust's novelty though is in \emph{ownership and borrowing}.
In Rust, each value is uniquely owned by a single variable, and
assigning that value to another transfers ownership of, and access to,
that value to the receiver. Rust calls this transfer of ownership a \emph{move}.
When a value is allowed to go out of scope
it is destroyed. In this way ownership-based resource management is
like RAII\footnote{``resource acquisition is initialization''} in C++.

To make using variables usably ergonomic Rust can create temporary
references to values with the ampersand (\lstinline[language=Rust]{&})
operator. References are simply pointers with limited scope, and may be either
immutable or mutable. They are the only type of pointer in safe Rust.
\emph{The rules for borrowing ensure that no two pointers are ever mutably
aliased, and that invalid memory is never accessible.
This is what makes Rust memory safe.} See the example in \figref{fig:own}.

@@

// Move the value in a to a new location
let moved_val = val;
// `val` is no longer accessible.
// The following will not compile.
// println(``{}'', val);

>>>>

Rust's most prominent influence from the ML languages are algebraic data types,
along with \emph{pattern matching} to destructure
them into their components. Rust calls algebraic data types \emph{enums},
and they are the union of multiple types,
each of which contains their own fields. Although superficially
similar to C unions, they are significantly different in that
instances of enums cannot be indiscriminately cast between
variants.
To access the values in an enum one must employ the \lstinline[language=Rust]{match}
expression to first check the variant, then bind references
to its interior fields. In \figref{fig:enums} the values of
a drawing message are extracted from an enum for further processing,
with the bindings to the enum's fields established to the left of the
fat arrow (\lstinline[language=Rust]{=>}) and the block of code operating
on those bindings to the right.

@@

Both enums and structs (which are substantially similar to C structs)
may have associated instance methods and static methods, called with the dot
(\lstinline[language=Rust]{.}) operator and the double-colon
(\lstinline[language=Rust]{::}) operator, respectively.
Static methods are common in Rust as they are used to construct
values, conventionally with a method called
\lstinline[language=Rust]{new}, as in
\lstinline[language=Rust]{Point2D::new(0.0, 0.0)}.
Notably, methods in Rust are dispatched statically, making them
eligible for inlining and aggressive optimization.
Methods are never dispatched dynamically through a function pointer in Rust
unless explicitly requested.

@@

Finally, Rust has a hygienic macro system, much more powerful than
the C preprocessor. Macro invocations look like function invocations
where the function name is appended with a bang (\lstinline[language=Rust]{!}).
They are not functions, but are instead expanded in place at compile time,
and perform syntactic transformations unavailable to functions.
The common method of performing console output, \lstinline[language=Rust]~println!("{}", foo)~,
is a macro.
See \figref{fig:bad-concurrency} for an example of closures and macros.

\subsection{Ownership}
The features of Rust described so far are common to many languages.
Rust's novelty is in \emph{ownership and borrowing}.
In Rust, each value is uniquely owned by a single variable, and
assigning that value to another variable transfers ownership of, and access to,
that value to the receiver. Rust calls this transfer of ownership a \emph{move}.
When a value is allowed to go out of scope
it is destroyed. In this way ownership-based resource management is
like RAII\footnote{``resource acquisition is initialization''} in C++.

To make using variables ergonomic, Rust can create temporary
references to values with the ampersand (\lstinline[language=Rust]{&})
operator. References are simply pointers with limited scope, and may be either
immutable or mutable. They are the only type of pointer in safe Rust.
\emph{The rules for borrowing ensure that no two pointers are ever mutably
aliased, and that invalid memory is never accessible.
This is what makes Rust memory safe.} See the example in \figref{fig:own}.

@@

// Move the value in `val` to a new location
let moved_val = val;
// `val` is no longer accessible.
// The following will not compile.
// println(``{}'', val);
